Descriptive Complexity for Counting Complexity Classes

### 2017-10-04 LMCS/proof-capture-fp.tex
[journal] cambios menores en demos

In this proof and the followings, we will reuse the symbol $<$ to denote the lexicographic order over same-sized tuples. Formally, for $\bar{x} = (x_1,\ldots,x_m)$ and $\bar{y} = (y_1,\ldots,y_m)$ we denote by $\bar{x} < \bar{y}$ the formula:
$$
\bigvee_{i = 1}^m \, \bigwedge_{j = 1}^{i-1} \, (x_j = y_j \wedge x_i < y_i).
$$
Similarly, we use $\bar{x} = \bar{y}$ to denote equality between tuples and $\bar{x} \leq \bar{y}$  to denote $\bar{x} < \bar{y} \vee \bar{x} = \bar{y}$.
%Furthermore, If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$.
We will also use some syntactic sugar in $\qso$ to simplify formulas. Specifically, we will use the {\em conditional count} symbol $(\varphi \mapsto \alpha)$ defined as $ (\varphi\cdot\alpha) + \neg\varphi$ for any Boolean formula $\varphi$ and any quantitative formula $\alpha$. Note that for each $\A \in \ostr[\R]$, and each first-order (second-order) assignment $v$ ($V$) over $\A$:
\[
\sem{(\varphi \mapsto \alpha)}(\A,v,V) = 
\begin{cases}
\sem{\alpha}(\A,v,V) &\text{if } (\A,v,V)\models\varphi,\\
1 &\text{otherwise}.
\end{cases}
\]

Now we prove Theorem~\ref{theo:capture-fp}. For condition (1), recall that checking whether $\A\models\varphi$ for any $\lfp$ formula $\varphi$ can be done in deterministic polynomial time on the size of $\A$\cite{I83}. 
%Also, the constant function $s$ can be trivially simulated in $\fp$. 
Furthermore, it is easy to check that $\fp$ is closed under polynomial sum and multiplication. 
We conclude then that any formula in $\qfo(\lfp)$ can be computed in $\fp$.	
For condition~(2), let $\R$ be a signature, $f\in \fp$
and $\ell\in\nat$ such that $\log_2\left(f(\enc(\A)) \right) \leq \size{\A}^\ell$ for every $\A\in\ostr[\R]$ (i.e. $\size{\A}^\ell$ is an upper bound for the output size of $f$ over $\A$).
Consider the language:
\[
L = \{(\A,\bar{a})\mid \bar{a} \in A^l \text{ and the } \bar{a}\text{-th bit of }f(\enc(\A))\text{ is 1}\}.
\]
where $\bar{a}$ encodes a number by following the lexicographic order over $A^l$.
%Consider a procedure that receives $\enc(\A)$ and an assignment $\bar{a}$ to $\bar{x}$. Let $m$ be the position of $\bar{a}$ in the lexicographic order of the tuples in $A^{\ell}$. The procedure then computes the $m$-th bit of $f(\enc(\A))$, from least to most significant. 
Clearly, the language $L$ is in $\ptime$ and by \cite{I83} there exists a formula $\Phi(\bar{x})$ in $\lfp$ such that $\A\models\Phi(\bar{a})$ if, and only if, $(\A,\bar{a})\in L$. 
We use then the following formula to encode $f$:
$$
\alpha = \sa{\bar{x}} \Phi(\bar{x})\cdot \pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2)$$
Note that the subformula $\pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2$ takes the value $2^m$ if there exist $m$ tuples in $A^{\ell}$ that are smaller than $\bar{x}$. Summing these values for each $\bar{a}\in A^{\ell}$ gives exactly $f(\enc(\A))$. 
In other words, $\Phi(\bar{x})$ simulates the behavior of the $\fp$-machine and the formula $\alpha$ reconstructs the binary output bit by bit.
Then $\alpha$ is in $\qfo(\lfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

>>>>

In this and the following proofs, we will reuse the symbol $<$ to denote the lexicographic order over same-sized tuples. 
Formally, for $\bar{x} = (x_1,\ldots,x_m)$ and $\bar{y} = (y_1,\ldots,y_m)$ we denote by $\bar{x} < \bar{y}$ the formula:
$$
\bigvee_{i = 1}^m \, \bigwedge_{j = 1}^{i-1} \, (x_j = y_j \wedge x_i < y_i).
$$
Similarly, we use $\bar{x} = \bar{y}$ to denote equality between tuples and $\bar{x} \leq \bar{y}$  to denote $\bar{x} < \bar{y} \vee \bar{x} = \bar{y}$.
%Furthermore, If $\bar{x} = (x_1,\ldots,x_m)$ ($\bar{X} = (X_1,\ldots,X_m)$) is a tuple of first-order (second-order) variables, we denote $\sa{\bar{x}}\alpha$ for $\sa{x_1}\cdots\sa{x_m}\alpha$ and $\sa{\bar{X}}\alpha$ for $\sa{X_1}\cdots\sa{X_m}\alpha$ for each $\qso$ formula $\alpha$.
We will also use some syntactic sugar in $\qso$ to simplify formulas. 
Specifically, we will use the {\em conditional count} symbol $(\varphi \mapsto \alpha)$ defined as $ (\varphi\cdot\alpha) + \neg\varphi$ for any Boolean formula $\varphi$ and any quantitative formula $\alpha$. 
Note that for each $\A \in \ostr[\R]$, and each first-order (second-order) assignment $v$ ($V$) over $\A$:
\[
\sem{(\varphi \mapsto \alpha)}(\A,v,V) = 
\begin{cases}
\sem{\alpha}(\A,v,V) &\text{if } (\A,v,V)\models\varphi,\\
1 &\text{otherwise}.
\end{cases}
\]
Furthermore, we use $\size{\A}$ to denote the size of an $\R$-structure $\A$.
Now we prove Theorem~\ref{theo:capture-fp}. 
For condition (1), recall that checking whether $\A\models\varphi$ for any $\lfp$ formula $\varphi$ can be done in deterministic polynomial time on the size of $\A$\cite{I83}. 
%Also, the constant function $s$ can be trivially simulated in $\fp$. 
Furthermore, it is easy to check that $\fp$ is closed under polynomial sum and multiplication. 
We conclude then that any formula in $\qfo(\lfp)$ can be computed in $\fp$.	
For condition~(2), let $\R$ be a signature, $f\in \fp$
and $\ell\in\nat$ such that $\log_2\left(f(\enc(\A)) \right) \leq \size{\A}^\ell$ for every $\A\in\ostr[\R]$ (i.e. $\size{\A}^\ell$ is an upper bound for the output size of $f$ over $\A$).
Consider the language:
\[
L = \{(\A,\bar{a})\mid \bar{a} \in A^l \text{ and the } \bar{a}\text{-th bit of }f(\enc(\A))\text{ is 1}\}.
\]
where $\bar{a}$ encodes a number by following the lexicographic order over $A^l$.
%Consider a procedure that receives $\enc(\A)$ and an assignment $\bar{a}$ to $\bar{x}$. Let $m$ be the position of $\bar{a}$ in the lexicographic order of the tuples in $A^{\ell}$. The procedure then computes the $m$-th bit of $f(\enc(\A))$, from least to most significant. 
Clearly, the language $L$ is in $\ptime$ and by \cite{I83} there exists a formula $\Phi(\bar{x})$ in $\lfp$ such that $\A\models\Phi(\bar{a})$ if, and only if, $(\A,\bar{a})\in L$. 
We use then the following formula to encode $f$:
$$
\alpha = \sa{\bar{x}} \Phi(\bar{x})\cdot \pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2)$$
Note that the subformula $\pa{\bar{y}}(\bar{y} < \bar{x}) \mapsto 2$ takes the value $2^m$ if there exist $m$ tuples in $A^{\ell}$ that are smaller than $\bar{x}$. Adding these values for each $\bar{a}\in A^{\ell}$ gives exactly $f(\enc(\A))$. 
In other words, $\Phi(\bar{x})$ simulates the behavior of the $\fp$-machine and the formula $\alpha$ reconstructs the binary output bit by bit.
Then $\alpha$ is in $\qfo(\lfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

### 2017-10-04 LMCS/proof-prop-rest.tex
[journal] cambios menores en demos

We give this proof in three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. Towards a contradiction, let $\R = \{ < \}$ and suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ which is equivalent to the $\E{1}$ formula $\sa{x} \ex{y} (x < y)$. This is, for every finite $\R$-structure $\A$ with $n$ elements, $\sem{\alpha}(\A) = n - 1$. We will use the following claim.
\begin{clm}
	Let $\alpha = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. Then $\alpha$ defines a function that is either null, greater or equal to $n$, or is in $\Omega(n^2)$.
\end{clm}
\proof
	Note that each atomic sub-formula in $\varphi(\bar{x})$ has either the form $(x = y)$, $(x < y)$, $\top$ or a negation thereof, where $x,y\in\bar{x}$. Suppose $\sem{\alpha}$ is not null and consider some $\R$-structure $\A$ such that $\sem{\alpha}(\A) > 0$. Let $\bar{a}$ be an assignment to $\bar{x}$ such that $\A\models\varphi(\bar{a})$. It can be seen that each assignment $\bar{a}'$ that has the same order over its variables\footnote{For example, if $\bar{a} = (a_1,a_2,a_3,a_4)$, such order may be $a_1 > a_3 = a_4 > a_2$.} as $\bar{a}$ also satisfies $\A\models\varphi(\bar{a}')$. If this order has $k$ partitions then $\binom{n}{k}$ assignments for $\bar{x}$ satisfy this order, and therefore, $\sem{\alpha} \geq \binom{n}{k}$. If $k = 1$, then $\binom{n}{k} = n$, and if $k \geq 2$, then $\binom{n}{k} \in \Omega(n^2)$, which proves the claim.
\qed
To prove that $\alpha$ does not define the $(n-1)$-function, suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fix $k$. Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. Note that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^n)$, as it was proven in~\cite{SalujaST95}. Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$ and, as we proved in the claim above, it describes either some function greater or equal to $n$, or in $\Omega(n^2)$, which leads to a contradiction.

@@

Finally, we prove that $\eqso(\loge{1})\subsetneq\eqso(\logu{1})$. For inclusion, let $\alpha$ be a formula in $\eqso(\loge{1})$. Suppose that it is in $\loge{1}$-SNF, namely, $\alpha = c + \sum_{i = 1}^{n}\alpha_i$. Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$, where $\varphi_i$ is quantifier-free for each $\alpha_i$. We use the same construction used in \cite{SalujaST95}, and we obtain that the formula $\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$ is equivalent to $\sa{\bar{y}}\,\varphi_i(\bar{X},\bar{x},\bar{y}) \wedge \fa{\bar{y}'}(\varphi_i(\bar{X},\bar{x},\bar{y}')\to\bar{y}\leq\bar{y}')$ for every assignment to $(\bar{X},\bar{x})$. We do this replacement for each $\alpha_i$ and we obtain an equivalent formula in $\eqso(\logu{1})$.

To prove that the inclusion is proper, consider the $\eqso(\logu{1})$ formula $\sa{x} \fa{y}(y = x)$. This formula defines the following function over each ordered structure $\A$:
$$
\sem{\alpha}(\A) = 
\begin{cases}
1 &\A \text{ has one element}\\
0 &\text{ otherwise}.
\end{cases}
$$
Suppose that there exists an equivalent formula $\alpha$ in $\eqso(\loge{1})$. Also, suppose that it is in $\L$-PNF, so $\alpha = c + \sum_{i = 1}^n\sa{\bar{X}}\sa{\bar{x}}\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$. Since $\alpha$ takes the value 0 for some structures, $c$ must be 0. Consider a structure $\A'$ with one element. We have that for some $i$, there exists an assignment $(\bar{B},\bar{b},\bar{a})$ for $(\bar{X},\bar{x},\bar{y})$ such that $\A' \models\varphi_i(\bar{B},\bar{b},\bar{a})$. Consider now the structure $\A''$ that is obtained by duplicating $\A'$, as we did for Theorem \ref{theo-pi1-pnf}. Note that $\A''\models\varphi_i(\bar{B},\bar{b},\bar{a})$, which implies that $\sem{\alpha}(\A' \uplus \A'') > 1$, which leads to a contradiction.

>>>>

We give this proof in three parts.
First, we show that $\E{1} \not\subseteq \QE{0}$. 
Towards a contradiction, let $\R = \{ < \}$ and suppose that there is a $\QE{0}$ formula $\alpha$ over $\R$ which is equivalent to the $\E{1}$ formula $\sa{x} \ex{y} (x < y)$. 
This is, for every finite $\R$-structure $\A$, $\sem{\alpha}(\A) = \size{\A} - 1$.

Suppose that $\alpha$ is in SNF, namely, $\alpha = \sum_{i = 1}^k \alpha_i$ for some fixed $k$. 
Since $\alpha$ is not null, consider some $\alpha_i$ that describes a non-null function. 
Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\varphi(\bar{X},\bar{x})$ where $\varphi$ is quantifier-free. 
Note that if $\length{\bar{X}} > 0$, then the function $\sem{\alpha}$ is in $\Omega(2^{\size{\A}})$, as it was proven in~\cite{SalujaST95}. 
Therefore, we have that $\alpha_i = \sa{\bar{x}}\varphi(\bar{x})$. 
We conclude our proof with the following claim.
\begin{clm}
	Let $\alpha = \sa{\bar{x}}\varphi(\bar{x})$	where $\varphi$ is quantifier free. 
	Then the function $\sem{\alpha}$ is either null, greater or equal to $n$, or is in $\Omega(n^2)$, where $n$ is the size of the input structure.
\end{clm}
\proof
	Note that each atomic sub-formula in $\varphi(\bar{x})$ is either $(x = y)$, $(x < y)$, $\top$ or a negation thereof, where $x,y\in\bar{x}$. 
	Suppose $\sem{\alpha}$ is not null and consider some $\R$-structure $\A$ such that $\sem{\alpha}(\A) > 0$. 
	Let $\bar{a}$ be an assignment to $\bar{x}$ such that $\A\models\varphi(\bar{a})$. 
	It can be seen that each assignment $\bar{a}'$ that has the same order over its variables\footnote{For example, if $\bar{a} = (a_1,a_2,a_3,a_4)$, such order may be $a_1 > a_3 = a_4 > a_2$, which has tree partitions.}	as $\bar{a}$ also satisfies $\A\models\varphi(\bar{a}')$. 
	If this order has $k$ partitions then $\binom{\size{\A}}{k}$ assignments for $\bar{x}$ satisfy this order, and therefore, $\sem{\alpha} \geq \binom{\size{\A}}{k}$. 
	If $k = 1$, then $\binom{\size{\A}}{k} = \size{\A}$, and if $k \geq 2$, then $\binom{\size{\A}}{k} \in \Omega(\size{\A}^2)$, which proves the claim.
\qed

@@

Finally, we prove that $\eqso(\loge{1})\subsetneq\eqso(\logu{1})$. 
For inclusion, let $\alpha$ be a formula in $\eqso(\loge{1})$. 
Suppose that it is in $\loge{1}$-SNF, namely, $\alpha = c + \sum_{i = 1}^{n}\alpha_i$. 
Let $\alpha_i = \sa{\bar{X}}\sa{\bar{x}}\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$, where $\varphi_i$ is quantifier-free for each $\alpha_i$. 
We use the same construction used in \cite{SalujaST95}, and we obtain that the formula $\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$ is equivalent to $\sa{\bar{y}}\,[\varphi_i(\bar{X},\bar{x},\bar{y}) \wedge \fa{\bar{y}'}(\varphi_i(\bar{X},\bar{x},\bar{y}')\to\bar{y}\leq\bar{y}')]$ for every assignment to $(\bar{X},\bar{x})$. 
Doing this replacement for each $\alpha_i$ renders an equivalent formula in $\eqso(\logu{1})$.

To prove that the inclusion is proper, consider the $\eqso(\logu{1})$ formula $\sa{x} \fa{y}(y = x)$. 
This formula defines the following function over each ordered structure $\A$:
$$
\sem{\alpha}(\A) = 
\begin{cases}
1 &\A \text{ has one element}\\
0 &\text{ otherwise}.
\end{cases}
$$
Suppose that there exists an equivalent formula $\alpha$ in $\eqso(\loge{1})$. 
Also, suppose that it is in $\L$-PNF, so $\alpha = \sum_{i = 1}^n\sa{\bar{X}}\sa{\bar{x}}\ex{\bar{y}}\varphi_i(\bar{X},\bar{x},\bar{y})$. 
Consider a structure $\A'$ with one element. 
We have that for some $i$, there exists an assignment $(\bar{B},\bar{b},\bar{a})$ for $(\bar{X},\bar{x},\bar{y})$ such that $\A' \models\varphi_i(\bar{B},\bar{b},\bar{a})$. 
Consider now the structure $\A''$ that is obtained by duplicating $\A'$, as we did for Theorem \ref{theo-pi1-pnf}. 
Note that $\A''\models\varphi_i(\bar{B},\bar{b},\bar{a})$, which implies that $\sem{\alpha}(\A' \uplus \A'') > 1$, which leads to a contradiction.
